---
title: 'Biogram: a toolkit for n-gram analysis'
author: "Michał Burdukiewicz, Piotr Sobczyk, Małgorzata Kotulska, Paweł Mackiewicz"
date: "02-06-2016"
output:
  ioslides_presentation: null
  fig_height: 5.5
  beamer_presentation: default
  slidy_presentation: default
fig_width: 9.5
widescreen: yes
---
  
```{r, echo = FALSE, message = FALSE, results='asis',warning=FALSE}
library(DT)
library(knitr)
library(ggplot2)
library(grid)
library(gridExtra)
library(dplyr)
library(biogram)
library(reshape2)
library(plotly)

options(DT.options = list(iDisplayLength = 6, searching = TRUE))

size_mod <- -4

enc2df <- function(x)
  data.frame(Names = names(x), Elements = toupper(sapply(x, paste0, collapse = ", ")))

my_theme <- theme(
  axis.text = element_text(size=13 + size_mod),
  axis.title.x = element_text(size=14 + size_mod, vjust = -1),
  axis.title.y = element_text(size=14 + size_mod, vjust = 1),
  
  legend.background = element_rect(fill = "NA"),
  legend.key = element_rect(fill = "NA", color = "NA", size = 0.5),
  legend.position = "bottom",
  #uncomment for publications
  legend.key.size = unit(0.1, "inches"),
  legend.margin = unit(-0.25, "lines"),
  legend.text = element_text(size=13 + size_mod), 
  legend.title = element_text(size=15 + size_mod),
  
  panel.grid.major = element_line(color="grey", linetype = "dashed", size = 0.5),
  panel.grid.major = element_line(color="lightgrey", 
                                  linetype = "dashed", size = 0.5),
  panel.background = element_rect(fill = "transparent", color = "black"),
  
  plot.background=element_rect(fill = "transparent",
                               color = "transparent"),
  #uncomment for publications
  plot.margin = unit(rep(0.02, 4), "inches"),
  plot.title = element_text(size=20 + size_mod),
  
  strip.background = element_rect(fill = "NA", color = "NA"),
  strip.text = element_text(size=13 + size_mod, face = "bold")
)
```

## Scientific problem

- How to predict properties of proteins using only amino acid sequences?  

## Amyloids

Amyloids are proteins associated with the number of clinical disorders (e.g. Alzheimer's or Creutzfeldt-Jakob's diseases) characterized by the occurence of hot-spots, short (6-15 amino acid) subsequences.  

Through hot-spots amyloids can create harmful zipper-like $\beta$-structures.

## n-grams

```{r, echo = FALSE, message = FALSE, results='asis',warning=FALSE}
sample_seq <- matrix(sample(c("A", "C", "G", "T"), 18, replace = TRUE), nrow = 3)
colnames(sample_seq) <- paste0("P", 1L:ncol(sample_seq))
rownames(sample_seq) <- paste0("S", 1L:nrow(sample_seq))

kable(data.frame(sample_seq), caption = "Sample sequences.  S - sequence, P - position.", digits = 0)
```

## n-grams

```{r, echo = FALSE, message = FALSE, results='asis',warning=FALSE}
unis <- count_ngrams(sample_seq, 1, c("A", "C", "G", "T"))
unis <- data.frame(as.matrix(unis))
colnames(unis) <- decode_ngrams(colnames(unis))
rownames(unis) <- paste0("S", 1L:nrow(sample_seq))
kable(unis, caption = "Counts of 1-grams.", digits = 0)
```

## n-grams

```{r, echo = FALSE, message = FALSE, results='asis',warning=FALSE}
unis <- count_ngrams(sample_seq, 2, c("A", "C", "G", "T"))
unis <- data.frame(as.matrix(unis))
colnames(unis) <- decode_ngrams(colnames(unis))
rownames(unis) <- paste0("S", 1L:nrow(sample_seq))
kable(unis, caption = "Counts of 2-grams.", digits = 0)
```

## n-grams

```{r, echo = FALSE, message = FALSE, results='asis',warning=FALSE}
unis <- count_ngrams(sample_seq, 2, d = 1, c("A", "C", "G", "T"))
unis <- data.frame(as.matrix(unis))
colnames(unis) <- sub("_", "|", decode_ngrams(colnames(unis)))
rownames(unis) <- paste0("S", 1L:nrow(sample_seq))
kable(unis, caption = "Counts of 2-grams with distance 1.", digits = 0)
```


## Encoding of amino acids

We created 524 284 reduced amino acid alphabets (encodings) with different lengths (from three to six letters) using Ward's clusterization on the selected physicochemical properties from AAIndex database.


## Best-performing encoding

```{r, echo = FALSE, message = FALSE, results='asis',warning=FALSE}
datatable(enc2df(list(`1` = "g", 
                  `2` = c("k", "p", "r"), 
                  `3` = c("i", "l", "v"), 
                  `4` = c("f", "w", "y"), 
                  `5` = c("a", "c", "h", "m"), 
                  `6` = c("d", "e", "n", "q", "s", "t"))), 
          rownames = FALSE, filter = "none", 
          options = list(dom = 't')) %>% 
  formatStyle(
    'Elements',
    backgroundColor = styleEqual("K, P, R", 'aquamarine')
  )
```

## Important n-grams

```{r, echo = FALSE, message = FALSE, results='asis',warning=FALSE,fig.width=10}
ngram_freq <- read.csv("ngram_freq.csv")

ngram_freq_plot <- mutate(ngram_freq, decoded_name = gsub("_", "|", decoded_name)) %>%
  mutate(decoded_name = factor(decoded_name, levels = as.character(decoded_name)),
         amyloid = diff_freq > 0) %>%
  melt() %>%
  filter(variable %in% c("pos", "neg")) %>%
  droplevels %>%
  mutate(variable = factor(variable, labels = c("Amyloid", "Non-amyloid"))) %>% 
  filter(association != "Not found") %>% 
  rename(Frequency = value) %>% 
  droplevels

ngram_plot <- ggplot(ngram_freq_plot, aes(x = decoded_name, y = Frequency)) +
  geom_bar(aes(fill = variable), position = "dodge", stat = "identity") +
  geom_point(data = group_by(ngram_freq_plot, decoded_name)  %>% filter(Frequency == max(Frequency)),
             aes(y = Frequency + 0.002, shape = association)) +
  scale_fill_manual("", values = c("gold", "darkmagenta")) +
  scale_shape_manual("Motif:", breaks = c("Amyloidogenic", "Non-amyloidogenic"), values = c(16, 17, NA)) +
  scale_y_continuous("Frequency") +
  scale_x_discrete("") +
  my_theme +
  theme(panel.grid.major.y = element_line(color = "lightgrey", size = 0.5)) 

ggplotly(ngram_plot, tooltip = "y")
```
